Investigating the outer reaches of the Sculptor and Ursa Minor dwarf galaxies

### 2025-07-23 thesis/introduction.tex
refining intro

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These little galaxies are typically
defined to fainter than the LMC or SMC \citep[\(M_V \gtrsim -18\),
e.g.][\citet{hodge1971}]{mcconnachie2012}. As the smallest class of
galaxies, dwarfs are the most numerous galaxies. Dwarf galaxies are
highly \emph{dark-matter dominated}, with mass to light ratios sometimes
exceeding 100 to 1000 \(M_\odot/ L_\odot\). Because of their small total
masses, many dwarf galaxies are \emph{quenched}, with little to no
recent star formation. Their stellar populations are \emph{relics} from
the early universe, consisting of many of the oldest and most metal poor
stars. Understanding the properties of dwarf galaxies thus has
implications across astronomy, from cosmological structure formation on
the smallest scale to extremely metal-poor stellar populations to
chemical evolution to galactic dynamics. Of the classical dwarfs, the
Sculptor and Ursa Minor dwarf spheroidal galaxies stand out with a more
extended density profile relative to an exponential, hinting at tidal
effects. As a case study in tidal effects and our understanding of dwarf
galaxy formation and evolution, we aim to understand the dynamical
history of these galaxies.

@@

Dark matter halos, and their associated galaxies, rarely evolve in
isolation. Instead, structure formation is \emph{hierarchical}. Small
dark matter halos collapse first and hierarchically merge into
progressively larger halos \citep[e.g.][\citet{blumenthal+1986},
\citet{white+rees1978}, \citet{white+frenk1991},
\citet{somerville+dave2015}]{blumenthal+1984}. Structure formation
happens on a wide range of scales. The largest structures become the
cosmic web---composed of voids, filaments, and clusters---and the
smallest structures directly observable host dwarf galaxies.
Hierarchical assembly is evident through the large scale structure of
the universe, remnants of past mergers within the milky way, and tidal
disruption of dwarf galaxies and their streams around nearby galaxies.

Small-scale structure formation is sensitive to deviations from
\(\Lambda\)CDM cosmology \citep[e.g.][]{bechtol+2022}. Many alternative
models, such as warm dark matter or self-interacting dark matter, may
smooth out small-scale features and reduce the abundance of small halos
or change their structure \citep[e.g.][]{lovell+2014}. Dwarf galaxies,
occupying the smallest dark matter halos, are promising windows into
small-scale cosmological features. Alternative dark matter properties
can influence dwarf galaxy structure, formation, and tidal evolution.
The nearby dwarf galaxies of the local group, with detailed
observations, present a promising opportunity to understand the
evolution of these objects and test our understanding of cosmology. To
understand the evolution of these objects, we need to understand the
general predictions from \(\Lambda\)CDM for the properties of dwarf
galaxies.

\subsection{Cosmological dark matter density
profiles}\label{cosmological-dark-matter-density-profiles}

In \(\Lambda\)CDM cosmological simulations, dark matter halos are
remarkably self-similar. In \citet{NFW1996};\citet{NFW1997}, hereafter
NFW, the authors observe that the spherical radial density profiles
\(\rho(r)\) are well described by a two-parameter law: \begin{equation}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density . This profile has shown remarkable success in describing
\(\Lambda\)CMD halos across several orders of magnitude in mass. NFW
profiles are \emph{cuspy}, where the density continuously rises like
\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
density profile increases at \(r \sim r_s\) and at large radii, the
density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
shows an example of a NFW halo and a cored dark matter halo.

@@

While two free parameters characterize a NFW halo, \(M_{200}\) and
\(c\), these parameters are not independent. Smaller dark matter halos
often collapse earlier, when the universe was denser. As a result, small
subhalos tend to be more concentrated \citep[e.g.][]{NFW1997}. The
relationship between \(M_{200}\) and c, or the mass-concentration
relation, how more massive halos become less concentrated in a
predictable way \citep[e.g.][\citet{ludlow+2016}]{bullock+2001}. While
there is a general trend in concentration with mass, the concentration
values still have substantial scatter. Other parameters such as the halo
spin or shape or slight changes in the halo density profile add
additional variability to the present-day distribution of halos
\citep[see e.g.][\citet{dutton+maccio2014}, darkEXP,
ect.]{navarro+2010}.

@@

\subsection{Connecting stars and dark
matter}\label{connecting-stars-and-dark-matter}

@@

Several challenges complicate a simple SMHM trend including environment,
assembly history, and tidal effects. By being closer to a massive host,
most dwarfs quench earlier, resulting in lower stellar masses
\citep[e.g.][]{christensen+2024}. Additionally, the time of formation
(relative to reionization) can influence the resulting stellar content
\citep{kim+2024}. Finally, tides influence both the dark matter and
stellar mass but in different amounts \citep[e.g.][]{PNM2008}.
Consequently, tides may reduce the halo mass more than the stellar mass,
adding additional scatter to the SMHM trend, particularly for Milky Way
satellites {[}@.e.g fattahi+2018{]}. Understanding the effects of tides
on local group dwarf galaxies is essential to determining where and how
these galaxies formed in a cosmological context.

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715100059332.png}
\caption[Stellar-mass halo-mass relation]{The stellar mass halo mass
relation, where \(v_{\rm max}\) is related to the halo mass. Taken from
\citet{fattahi+2018}. Right: the mass-concentration relation for NFW
halos, but parameterized in terms of \(r_{\rm max}\) and
\(v_{\rm max}\). Together, these plots represent the cosmologically
expected properties of underlying dark matter halos in any dwarf galaxy.
The velocity dispersion directly constrains the mass contained within a
half-light radius, so the underlying halo in \(\Lambda\)CDM is
well-constrained.}\label{fig:smhm}
\end{figure}

@@

\subsection{Early observations}\label{early-observations}

Dwarf galaxies have often puzzled astronomers and our understanding of
the universe. While we now generally understand how dwarf galaxies fit
into the universe, this was not always the case. The first discovered
dwarf galaxies in 1938\footnote{technically, the LMC and SMC may be
  classified as dwarf galaxies, but these were likely always known to
  humans at southern latitudes.}, Sculptor and Fornax, heralds an era of
discovery and investigation into these astronomical system.
\citet{shapley1938} present Sculptor and Fornax as a new type of
\emph{stellar system} resembling the Magellanic clouds and globular
clusters, but do not attempt to speculate on the exact nature. The first
direct hints that these systems are galaxies originates from early
spectroscopic work deriving the first velocity dispersions
\citep[e.g.,][\citet{gallagher+wyse1994}, \citet{mateo1994},
\citet{pryor1994}, \citet{pryor1996}, \citet{gerhard1994}, and
\citet{olszewski1998}.]{aaronson1983}. Based on the physical size and
velocity dispersions, the mass to light ratios of these dwarf galaxies
is far larger than expected for star clusters without dark matter. In
general, these objects have reached a consensus that they have high
velocity dispersion, old spheroidal stellar populations, and low stellar
masses.

Dwarf galaxies span a large range of size and morphology.
Fig.~\ref{fig:galaxy_images} displays several images of dwarf galaxies.
The largest in the local group but not always considered a dwarf galaxy,
the Large Magellanic Cloud (LMC), to classical systems Sculptor and Ursa
Minor dwarf spheroidal, to an ultra-faint dwarf galaxy, Boötes V. Large
LMC-like dwarf galaxies exhibit complex structure such as a bar, young
blue stars, and irregular shapes. On the other hand, the classical and
faint dwarf galaxies are composed mostly of old, faint, and red stars.
These galaxies are typically classified as dwarf spheroidals (dSph)
(although we typically omit the dSph designation and refer to these
galaxies by their constellation alone).

Since the discovery of dwarf galaxies around the Milky Way,
observational work has attempted to measure and refine their basic
properties \citep[e.g.][]{mateo1998}. While the Milky Way's satellites
are close (by extragalactic standards), their low numbers of stars and
large apparent sizes present challenges for observational work. In
particular, the properties of the faintest dwarf galaxy systems and
especially \emph{ambiguous} systems remains a pressing challenge as a
unique probe into the nature of the early universe and dark matter.

\emph{Classical systems}, the brightest of the local group
(non-Magellanic) dwarf galaxies, are perhaps the systems best studied
and with the most stringently derived properties. Defined\footnote{Note
  that some work like \citet{simon2019} instead define classical dwarfs
  based on luminosity as compared to ultra faint dwarfs. This may
  include galaxies such as Antlia I, Crater II, Canes Venatici I, and
  Sagittarius. We stick with the historical definition for simplicity
  and since the Gaia observations are most numerous for the historical
  classicals.} as the first six discovered dwarf spheroidal galaxies,
(Sculptor, Ursa Minor, Fornax, Draco, Leo I, Leo II, and Carina),
classical systems are typically old, spheroidal, and gas-poor. While
extending 1-2 degrees across the sky (e.g.
Figs.~\ref{fig:scl_selection}, \ref{fig:umi_selection}, \ref{fig:fornax_selection}),
these dwarfs contain large numbers of bright (giant) stars, allowing
thousands of stars to be observed with deep photometry and spectroscopy
\citep[e.g.][\citet{pace+2020}]{tolstoy+2023}. As a result, the
determination of fundamental properties such as the position, size,
orientation, distance \citep[from 100s of RRL stars,
e.g.][]{tran+2022, garofalo+2025}, proper motions (from \emph{Gaia},
e.g. \citet{battaglia+2022}, \citet{MV2020a}), line-of-sight (LOS)
velocity and dispersion \(\sigma_v\), are all relatively well
constrained today and show excellent convergence among different works.
However, ongoing research continues to redefine our understanding of the
detailed structure of Milky Way satellites, hinting that these objects
may be more complex than at first glance.

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.pdf}
\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Sculptor (DES
2), Ursa Minor (UNWISE with Gaia point sources over-plotted), and Bootes
V (SDSS). While the LMC is very prominent in the sky, even the classical
dwarf galaxies are not obvious except in terms of star counts (e.g.
Fig.~\ref{fig:scl_selection}). All (non-LMC) dwarfs occupy the central
third to sixth of the image.}\label{fig:galaxy_images}
\end{figure}

@@

Some of the most fundamental properties of astronomical objects are
their positions and velocities. Unfortunately, determining distances to
stars is nontrivial. Additionally, while line-of-sight (LOS) velocities
are easily determined from spectroscopy, the tangental velocities,
perpendicular to the LOS, are best measured through proper motions,
typically requiring accurate determinations of small changes in a star's
position. \emph{Gaia}'s mission addresses these problems by making
extremely precise measurements of stellar positions over time.

\emph{Gaia} was designed to revolutionize proper motion and parallax
measurements. \emph{Gaia} is a space-based, all-sky survey telescope
with two primary 1.45x0.5m mirrors situated at the Sun-Earth L2 Lagrange
point \citep{gaiacollaboration+2016}. \emph{Gaia} was launched in 2013,
completing its space-based mission in 2025 (with two more data releases
planned). By imagining two patches of sky on the same focal plane,
separated by a fixed angle of 106.5 degrees, \emph{Gaia} is able to
measure absolute proper motions. Stars in different regions of the sky
are affected by parallax differently, so by precisely observing the
relative positions of stars separated by large angles during multiple
epochs over a year, an absolute all-sky reference frame can be derived
\citep{gaiacollaboration+2016}. In addition to precise astrometric
information, \emph{Gaia} measures the magnitude of stars in the very
wide \emph{G} band (330-1050nm), blue and red colours using the blue and
red photometers (BP and RP, 330-680 nm and 640-1050 nm respectively),
and takes low resolution BP-RP spectra and radial velocity measurements
of bright stars (\emph{Gaia} radial velocity magnitudes \textless16)
\citep{gaiacollaboration+2016}. For our purposes, most useful are
\emph{Gaia}'s measurements of \(G\) magnitude,
\(G_{\rm BP} - G_{\rm RP}\) colour, the position \(\alpha, \delta\), and
the proper motions in RA and declination \(\mu_{\alpha*}\) and
\(\mu_\delta\).\footnote{\(\mu_{\alpha*}\) allows where
  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
  effects{]}.}

\emph{Gaia} has revolutionized many astronomical disciplines, the least
of which is local group and Milky Way science. While proper motions of a
dwarf galaxies has been measured in a case by case basis by the Hubble
Space Telescope, a full systematic determination of proper motions for
most dwarf galaxies was unavailable until \emph{Gaia}
\citep{MV2020a, battaglia+2022}. \emph{Gaia} has furthermore allowed for
the detection and improved measurements of halo substructures and
streams \citep{bonaca+price-whelan2025}, Milky Way structure and
kinematics \citep{hunt+vasiliev2025}.

For Local group dwarf spheroidal, \emph{Gaia} has allowed for many of
the first proper motion measurements. While the proper motion
uncertainty on a typical dwarf member star is often with large
uncertainties, by combining the proper motions for 100s to 1000s of
stars in \emph{Gaia}, incredibly precise proper motion measurements can
be determined \citep[e.g.][]{MV2020a}, sometimes only limited by
\emph{Gaia}'s systematic error floor. Proper motions have ushered in a
new dynamical era for MW satellites studies, where we often have precise
determinations of the dwarf galaxy's (short term) orbit under a given MW
potential.

@@

With the help of \emph{Gaia}, high resolution spectroscopy, high-quality
photometry, and large aperture multi-object spectroscopic instruments,
we can observe dwarf galaxies in unprecedented detail. Today, dwarf
galaxies are used to search for dark matter cores (REFS), test our
understanding of early nucleosynthesis (REFS), the details of galaxy
formation (REFS), and \ldots.

While we now know the basic properties of dwarf galaxies, tremendous
progress has been made in the past 10-20 years with large area deep
photometric surveys, multi-object spectrographs, high resolution
chemistry, and theoretical work, unveiling that dwarf galaxies are each
unique systems with different formation pathways
\citep[e.g.][]{simon2019}.

\citet{battaglia+nipoti2022}

One of the open questions in dwarf galaxy evolution is the formation of
dark matter \emph{cores} where the density becomes constant instead
\(\rho \sim 1\). A leading theory is cores form through of baryonic
effects and feedback. If dark matter deviates from \(\Lambda\)CDM, then
the NFW density profile may not describe halos anymore. \emph{Warm} dark
matter, relativistic in the early universe but cooler now, smooths out
small-scale features and softens the cusps of dark matter halos.
\emph{Self-interacting} dark matter instead can form cores but also the
cores can collapse into a density peak.

\begin{itemize}
\tightlist
\item
  \citet{walker+penarrubia2011}: measuring cores and cusps
\item
  \citet{dicintio+2013}, \citet{navarro+2010}, NFW profiles may actually
  be more Einasto like
\end{itemize}

@@

Given star counts for a local group dwarf, one of the main quantities of
interest is the surface density profile. Long used to characterize and
understand larger galaxies, these profiles can provide hints as to the
formation of dwarf galaxies, determine the physical extent of the
stellar component, and the total stellar mass. Previous works tend to
focus on fitting 3 different surface brightness profiles: an
Exponential, a Plummer, or a King profile.

The exponential profile is perhaps the simplest, as a 1-parameter
profile \begin{equation}{
\Sigma_{\rm exp} = \Sigma_0\exp(-R / R_s)
}\end{equation}

For a long time, exponential surface density profiles have been used as
an description for spiral galaxies {[}REFS{]}.

\citet{plummer1911} proposed a 1-parameter solution for a polytropic
potential to fit globular cluster density profiles:

\begin{equation}{
\Sigma_{\rm Plummer} = \frac{M}{\pi R_h^2}\frac{1}{(1 + (R/R_h)^2)^2}
}\end{equation}

The \citet{king1962} profile, originating as an empirical fit to
globular clusters, also sometimes reasonably fits dwarf galaxies. This
density profile includes an additional parameter \(R_t\), a truncation
radius where the density sharply drops to 0. \begin{equation}{
\Sigma_{\rm King} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right)
}\end{equation}

Finally, the Sérsic profile represents an extension of the exponential.
Typically parameterized in terms of a half-light radius \(R_h\), the
density at half-light radius \(\Sigma_h\) and a Sérsic index \(n\), the
profile's equation is \begin{equation}{
\Sigma_{\rm Sérsic} = \Sigma_0 \exp\left[-b_n \,  \left((R/R_h)^{1/n} - 1\right)\right]
}\end{equation} where \(b_n\) is a constant depending on \(n\). Less
commonly used in studies of dwarf galaxies, \citet{munoz+2018} advocate
for using the Sérsic profile since the added flexibility allows for more
profiles to be fit.

>>>>

Dwarf galaxies host, in many ways, the most extreme galactic
environments in the universe. These little galaxies are typically
defined to fainter than the LMC or SMC \citep[\(M_V \gtrsim -18\) ,
e.g.,][]{hodge1971, mcconnachie2012}. As the smallest class of galaxies,
dwarfs are the most numerous galaxies. Dwarf galaxies are highly
\emph{dark-matter dominated}, with mass to light ratios sometimes
exceeding 100 to 1000 \(M_\odot/ L_\odot\). Because of their small total
masses, many dwarf galaxies are \emph{quenched}, with little to no
recent star formation. Their stellar populations are \emph{relics} from
the early universe, consisting of many of the oldest and most metal poor
stars. Understanding the properties of dwarf galaxies thus has
implications across astronomy, from cosmological structure formation on
the smallest scale to extremely metal-poor stellar populations to
chemical evolution to galactic dynamics. Of the classical dwarfs, the
Sculptor and Ursa Minor dwarf spheroidal galaxies stand out with a more
extended density profile relative to an exponential, hinting at tidal
effects. As a case study in tidal effects and our understanding of dwarf
galaxy formation and evolution, we aim to understand the dynamical
history of these galaxies.

@@

Dark matter halos, and their associated galaxies, rarely evolve in
isolation. Instead, structure formation is \emph{hierarchical}. Small
dark matter halos collapse first and hierarchically merge into
progressively larger halos
\citep[e.g.,][]{blumenthal+1984, blumenthal+1986, white+rees1978, white+frenk1991, somerville+dave2015}.
Structure formation happens on a wide range of scales. The largest
structures become the cosmic web---composed of voids, filaments, and
clusters---and the smallest structures directly observable host dwarf
galaxies. Hierarchical assembly is evident through the large scale
structure of the universe, remnants of past mergers within the milky
way, and tidal disruption of dwarf galaxies and their streams around
nearby galaxies.

Small-scale structure formation is sensitive to deviations from
\(\Lambda\)CDM cosmology \citep[e.g.,][]{bechtol+2022}. Many alternative
models, such as warm dark matter or self-interacting dark matter, may
smooth out small-scale features and reduce the abundance of small halos
or change their structure \citep[e.g.,][]{lovell+2014}. Dwarf galaxies,
occupying the smallest dark matter halos, are promising windows into
small-scale cosmological features. Alternative dark matter properties
can influence dwarf galaxy structure, formation, and tidal evolution.
The nearby dwarf galaxies of the local group, with detailed
observations, present a promising opportunity to understand the
evolution of these objects and test our understanding of cosmology. To
understand the evolution of these objects, we need to understand the
general predictions from \(\Lambda\)CDM for the properties of dwarf
galaxies.

\subsection{Typical structure of dark matter
halos}\label{typical-structure-of-dark-matter-halos}

In \(\Lambda\)CDM cosmological simulations, dark matter halos are
remarkably self-similar. In \citet{NFW1996, NFW1997}, hereafter NFW, the
authors observe that the spherical radial density profiles \(\rho(r)\)
are well described by a two-parameter law: \begin{equation}{
\rho/\rho_s= \frac{1}{(r/r_s)(1+r/r_s)^2},
}\end{equation} where \(r_s\) is a scale radius and \(\rho_s\) a scale
density . This profile has shown remarkable success in describing
\(\Lambda\)CMD halos across several orders of magnitude in mass. NFW
profiles are \emph{cuspy}, where the density continuously rises like
\(\rho \sim 1/r\) at small radii \(r \ll r_s\). The steepness of the
density profile increases at \(r \sim r_s\) and at large radii, the
density falls off like \(\rho \sim 1/r^3\). Fig.~\ref{fig:nfw_density}
shows an example of a NFW halo and a cored dark matter halo.

@@

While an NFW profile has two free parameters (\(M_{200}\) and \(c\)),
these are not independent. Smaller dark matter halos often collapse
earlier, when the universe was denser. As a result, small subhalos tend
to be more concentrated \citep[e.g.,][]{NFW1997}. The relationship
between \(M_{200}\) and c, or the mass-concentration relation, how more
massive halos become less concentrated in a predictable way
\citep[e.g.,][]{bullock+2001, ludlow+2016}. While there is a general
trend in concentration with mass, the concentration values still have
substantial scatter. Other parameters such as the halo spin or shape or
slight changes in the halo density profile add additional variability to
the present-day distribution of halos \citep[see
e.g.,][darkEXP]{navarro+2010, dicintio+2013, dutton+maccio2014}.

@@

\subsection{Connecting dark matter to
stars}\label{connecting-dark-matter-to-stars}

@@

Several challenges complicate a simple SMHM trend including environment,
assembly history, and tidal effects. By being closer to a massive host,
most dwarfs quench earlier, resulting in lower stellar masses
\citep[e.g.,][]{christensen+2024}. Additionally, the time of formation
(relative to reionization) can influence the resulting stellar content
\citep{kim+2024}. Finally, tides influence both the dark matter and
stellar mass but in different amounts \citep[e.g.,][]{PNM2008}.
Consequently, tides may reduce the halo mass more than the stellar mass,
adding additional scatter to the SMHM trend, particularly for Milky Way
satellites \citep[e.g.,][]{fattahi+2018}. Understanding the effects of
tides on local group dwarf galaxies is essential to determining where
and how these galaxies formed in a cosmological context.

\begin{figure}
\centering
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{/Users/daniel/Library/Application Support/typora-user-images/image-20250715100059332.png}
\caption[Stellar-mass halo-mass relation]{The stellar mass halo mass
relation, where \(v_{\rm max}\) is related to the halo mass. Right: the
mass-concentration relation for NFW halos, but parameterized in terms of
\(r_{\rm max}\) and \(v_{\rm max}\). Together, these plots represent the
cosmologically expected properties of underlying dark matter halos in
any dwarf galaxy. The velocity dispersion directly constrains the mass
contained within a half-light radius, so the underlying halo in
\(\Lambda\)CDM is well-constrained. Adapted from figure 1 of
\citet{fattahi+2018}, stellar mass and \(v_{\rm max}\) of APOSTLE
satellites at peak \(v_{\rm max}\).}\label{fig:smhm}
\end{figure}

@@

Dwarf galaxies have long challenged questioned theories of galaxy
formation. The first discovered dwarf galaxies in 1938\footnote{technically,
  the LMC and SMC may be classified as dwarf galaxies, but these were
  likely always known to humans at southern latitudes.}, Sculptor and
Fornax, already presented such a puzzle. Having no close analogues,
\citet{shapley1938} present Sculptor and Fornax as a new type of
\emph{stellar system} resembling the Magellanic clouds and globular
clusters, but not attempting to speculate on the exact nature. While
generally understood to be galaxies based on the inferred luminosities
and sizes, the nature and formation remained unclear
\citep[e.g.,][]{hodge1971}.

The earliest spectroscopic work hinted that dwarf galaxies may contain
substantial dark matter. From the first measurements of velocity
dispersions
\citep[e.g.,][]{aaronson1983, gallagher+wyse1994, pryor1996}, the
inferred mass-to-light ratios were about 10 times the expectation for
globular clusters \citep{matteo1998}. Subsequently, many theories
attempting to understand the formation and observed properties of these
objects were proposed. Examples included that dwarf galaxies were formed
tidally in the gas clouds of satellites; dwarf galaxies are undergoing
tidal dissolution resulting in extreme mass-to-light ratios
\citep[e.g.,][]{pryor1996}, or Since then, we have known that dwarf
galaxies are among the darkest objects in the universe, and
understanding their properties is critical to understanding the
universe.

Dwarf galaxies span a large range of physical sizes, luminosities, and
morphologies. Broadly, there are three classes of dwarf galaxies based
on luminosity, as illustrated by Fig.~\ref{fig:galaxy_images}. Local
\textbf{bright dwarfs galaxies} with magnitudes
\(-18 \lesssim M_V \lesssim  -14\), often exhibit irregular morphologies
and recent star formation. Fig.~\ref{fig:galaxy_images} shows the LMC as
an example of an irregular (and slightly bright) dwarf galaxy displaying
a bar. \textbf{Classical dwarf} galaxies occupy intermediate
luminosities ( \(-14 \lesssim M_V  \lesssim -7.7\)). Typically, these
systems are old, gas-poor, and spheroidal. All Milky Way satellites
discovered before digital sky surveys are classicals. The 12 classical
dwarfs satellites of our galaxy are Sagittarius, Fornax, Leo I,
Sculptor, Antlia II, Leo II, Carina, Draco, Ursa Minor, Canes Venatici
I, Sextans I, and Crater II.\footnote{While formally the dwarf galaxy
  names we discuss contain'' dwarf spheroidal'' (dSph), e.g.~Sculptor
  dSph, we omit this suffix for brevity.} The ultra-faints occupy the
very faintest magnitudes (\(-7.7 \lesssim M_V\)). These galaxies have
miniscule stellar masses, tend to be more compact, and are the darkest
known galaxies. An example is Boötes V as shown in
Fig.~\ref{fig:galaxy_images}. Altogether, dwarf galaxies span about 7
orders of magnitude in stellar mass.

\begin{figure}
\centering
\includegraphics[width=5.41667in,height=5.41667in]{figures/galaxy_pictures.pdf}
\caption[Dwarf Galaxy Pictures]{Images of the LMC (DSS2), Sculptor (DES
2), Ursa Minor (UNWISE with Gaia point sources over-plotted), and Bootes
V (SDSS). While the LMC is very prominent in the sky, even the classical
dwarf galaxies are not obvious except in terms of star counts
(e.g.,Fig.~\ref{fig:scl_selection}). All (non-LMC) dwarfs occupy the
central third to sixth of the image.}\label{fig:galaxy_images}
\end{figure}

@@

A star's or galaxy's position and velocity are fundamental quantities
for understanding its orbit and origin. Unfortunately, determining
distances and full 3D velocities is challenging. The most direct
measurement of distance, parallax, requires precise tracking of a star's
sky position across a year. And while line-of-sight (LOS) velocities are
easily determined from spectroscopy, tangental velocities, derived from
proper motions, are much more challenging. Typically, measuring proper
motions requires accurate (much less than arcsecond) determinations of
small changes in a star's position over baselines of several years to
decades. The full 6D position and velocity information for stars was
historically only known for a handful of stars.

\emph{Gaia} has redefined astrometry, providing photometry, positions,
proper motions, and parallaxes for over 1 billion stars
\citep{gaialcollaboration+2021}. Launched in 2013, \emph{Gaia} is a
space-based, all-sky survey telescope situated at the Sun-Earth L2
Lagrange point \citep{gaiacollaboration+2016}. While \emph{Gaia}
completed its space-based mission in 2025, two forthcoming data releases
remain. Determining absolute parallax measurement is facilitated by the
observation that stars in different regions of the sky are affected by
parallax motion with different phases. By imagining two regions
separated by 106.5 degrees on the same focal plane, \emph{Gaia} measures
tiny changes in relative positions of stars across small and large
angles. Combining measurements from multiple epochs across several
years, an absolute all-sky reference frame is derived from which
parallax and proper motions are derived \citep{gaiacollaboration+2016}.
In addition to astrometry, \emph{Gaia} measures photometry in the wide
\emph{G} band (330-1050nm) and colours from the blue photometer (BP,
330-680 nm) and red photometer (RP, 640-1050 nm). \emph{Gaia}
additionally provides low resolution BP-RP spectra and radial velocity
measurements of bright stars (\emph{Gaia} radial velocity magnitudes
\textless16) \citep{gaiacollaboration+2016}. For this work,
\emph{Gaia}'s most relevant measurement are \(G\) magnitude,
\(G_{\rm BP} - G_{\rm RP}\) colour, \((\alpha, \delta)\) position, and
\((\mu_{\alpha*}, \mu_\delta)\) proper motions.\footnote{\(\mu_\alpha\)
  and \(\mu_\delta\) are the apparent changes in \(\alpha\) and
  \(\delta\). \(\mu_{\alpha*}\) allows where
  \(\mu_{\alpha*} = \mu_\alpha \cos \delta\) corrects for projection
  effects in \(\alpha\).}

\emph{Gaia} has revolutionized many fields of astronomy including the
Milky Way and the Local Group. For example, 6D kinematic classification
of Milky Way stars has lead to the discovery of past mergers or Milky
Way building blocks like \emph{Gaia}-sausage Enceladus
\citep[e.g.,][]{helmi+2018}, out-of-equilibrium structures like the
\emph{Gaia} snail {[}e.g., \citet{antoja+2018l}, and effects of the
Milky Way's bulge and bar \citep{hunt+vasiliev2025}. Moving to the Milky
Way halo, numerous streams \citep{bonaca+price-whelan2025}. \emph{Gaia}
has revealed the hierarchical formation and messy structure of our own
galaxy.

For Milky Way satellites, \emph{Gaia} has enabled well-constrained
orbital analysis and facilitated precise stellar membership
determination. Before, proper motions of dwarf galaxies were spase or
undetermined and measured by the Hubble Space Telescope or other methods
for only a few systems. \emph{Gaia} allowed for the first systematic
determinations of Milky Way satellite proper motions, and of
unprecedented precision \citep{MV2020a, pace+li2019}. While the proper
motion uncertainty on a typical dwarf member star is often with large
uncertainties, by combining the proper motions for 100s to 1000s of
stars in \emph{Gaia}, incredibly precise proper motion measurements can
be determined, sometimes only limited by \emph{Gaia}'s systematic error
floor \citep[e.g.,][]{MV2020a}. Proper motions have ushered in a new
dynamical era for MW satellites studies, where we often have precise
determinations of the dwarf galaxy's recent orbit under a given MW
potential. In addition, \emph{Gaia} helps seperate out contaiminating MW
foreground stars. By measuring parallaxes or proper motions
significantly different from the dwarf galaxy, many stars can be
immediately removed as non-members
\citep[e.g.,][\citet{jensen2024}]{battaglia+2022}.

@@

Today, we know the Milky Way system is teeming with satellites.
Fig.~\ref{fig:mw_satellite_system} shows the MW satellite system,
including dwarf galaxies, ambiguous systems, and globular clusters.
Advances in telescopes and observations has accelerated progress on the
nature of dwarf galaxy and introduced new questions. Deep digital
photometric surveys have more than quadrupled the number of known Milky
Way dwarf galaxies \citep{simon2019}. Upcoming and ongoing surveys, like
the Vera Rubin Observatory's Legacy Survey for Space and Time, will
continue to probe fainter and fainter dwarf galaxies. In addition, large
aperture multi-object spectrographs have revealed the detailed and
complex inner chemodynamical structure of dwarf galaxies. Beyond precise
structural and kinematic properties of dwarf galaxies, modern
observations allow for the separation of multiple stellar populations,
detailed constraints on the dark matter density profiles, and hints of
tidal disruption or stellar halos.

Of local dwarfs, the classical systems still remain the best studied and
with most strongly constrained parameters. While extending 1-2 degrees
across the sky, making deriving properties trickier
\citep[e.g.,][]{mateo1998}, these dwarfs contain large numbers of bright
(giant) stars, allowing thousands of stars to be observed with deep
photometry and spectroscopy \citep[e.g.,][]{tolstoy+2023, pace+2020}. As
a result, the determination of fundamental properties such as the
position, size, orientation, distances, proper motions, line-of-sight
(LOS) velocities, and velocity dispersions \(\sigma_v\), are all
relatively well constrained today and show excellent convergence among
different works. However, ongoing research continues to redefine our
understanding of the detailed structure of Milky Way satellites, hinting
that these objects may be more complex than at first glance.

Observations of dwarf galaxies have been the origin of several
\emph{small-scale} problems in cosmology
\citep[e.g.,][]{bullock+boylan-kolchin2017}. For example, the mismatch
between the expected number of dwarf galaxies from simulations and the
observed abundance was known as the \emph{missing satellites problem}.

\emph{Core Cusp Problem}. One of the open questions in dwarf galaxy
evolution is the formation of dark matter \emph{cores} where the density
becomes constant instead \(\rho \sim 1\). A leading theory is cores form
through of baryonic effects and feedback. If dark matter deviates from
\(\Lambda\)CDM, then the NFW density profile may not describe halos
anymore. \emph{Warm} dark matter, relativistic in the early universe but
cooler now, smooths out small-scale features and softens the cusps of
dark matter halos. \emph{Self-interacting} dark matter instead can form
cores but also the cores can collapse into a density peak.

\begin{itemize}
\tightlist
\item
  \citet{walker+penarrubia2011}: measuring cores and cusps
\item
  \citet{oh+2011} survey of cusyp / cored dwarf galaxies.
\end{itemize}

Understanding the p

@@

Surface density profiles efficiently characterize the shape of a galaxy.
At the most basic, fitting stellar densities provide properties such as
the shape, location, size, and orientation of a dwarf galaxy. However,
the details of a stellar density profile are essential for interpreting
the total mass of dwarf galaxies, their assembly and dynamical history,
and understanding correlations between dwarf galaxies (REFS).

Typically, three different surface density profiles are applied to dwarf
galaxies: Exponential, Plummer, or a King profile
\citep[e.g.,][]{munoz+18, cicuendez+2018}. The exponential profile is
perhaps the simplest, as a 1-parameter profile \begin{equation}{
\Sigma_{\rm exp} = \Sigma_0\exp(-R / R_s)
}\end{equation}

For a long time, exponential surface density profiles have been used as
an description for spiral galaxies (see discussion before).

To fit globular cluster density profiles, \citet{plummer1911} proposed a
1-parameter solution for a density polytrope\footnote{where density and
  pressure are assumed to be correlated}, \begin{equation}{
\Sigma_{\rm Plummer} = \frac{M}{\pi R_h^2}\frac{1}{(1 + (R/R_h)^2)^2},
}\end{equation}

where \(M\) is the total mass and \(R_h\) is the 2D half-light radius.
Now superceded by the King profile for globular clusters, the Plummer
model is still a good fit to dwarf spheroidals and XXX.

The \citet{king1962} profile, also an empirical fit to globular
clusters, is also commonly applied to . This density profile includes an
additional parameter \(R_t\), a truncation radius where the density
sharply drops to 0. \begin{equation}{
\Sigma_{\rm King} = \Sigma_0\left(\frac{1}{\sqrt{1 + (R/R_c)^2}} - \frac{1}{\sqrt{1+(R_t/R_c)^2}}\right)
}\end{equation}

Where \(R_c\) is the characteristic ``core'' radius and \(\Sigma_0\) is
a characteristic density related to the total mass.

Finally, the Sérsic profile represents a generalization of an
exponential profile, and describes galaxies of many kinds well.
Typically parameterized in terms of a half-light radius \(R_h\), the
density at half-light radius \(\Sigma_h\) and a Sérsic index \(n\), the
profile's equation is \begin{equation}{
\Sigma_{\rm Sérsic} = \Sigma_0 \exp\left[-b_n \,  \left((R/R_h)^{1/n} - 1\right)\right]
}\end{equation} where \(b_n\) is a constant depending on \(n\). Less
commonly used in studies of dwarf galaxies, \citet{munoz+2018} advocate
for using the Sérsic profile since the added flexibility allows for more
profiles to be fit.
